---
title: "Survival Analysis of Breastfeeding Duration"
author: "Andrew Kamya"
date:  "November, 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
Survival analysis is a statistical approach used to analyze time-to-event data. In this project, we examine the duration of breastfeeding using the `bfeed` dataset from the `KMsurv` package. We apply non-parametric, semi-parametric, and parametric survival models to understand factors influencing breastfeeding duration.


# Data Preparation
We begin by loading the necessary packages and dataset.

```{r load-data}
library(survival)
library(KMsurv)
data(bfeed)
head(bfeed)
```

The key variables of interest are:
- `duration`: Breastfeeding duration in weeks.
- `delta`: Indicator for completed breastfeeding (1=yes, 0=no).
- `smoke`: Whether the mother smoked at the child's birth (1=yes, 0=no).

# Kaplan-Meier Estimation
We estimate and plot the survival functions for smoking and non-smoking mothers.

```{r kaplan-meier}
data1 <- subset(bfeed, smoke == 1)
data0 <- subset(bfeed, smoke == 0)

fit1 <- survfit(Surv(duration, delta) ~ 1, data = data1)
fit0 <- survfit(Surv(duration, delta) ~ 1, data = data0)

plot(fit0, col = "blue", conf.int = FALSE, main = "Kaplan-Meier Survival Curves",
     xlab = "Duration (weeks)", ylab = "Survival Probability")
lines(fit1, col = "red", conf.int = FALSE)
legend("topright", legend = c("Non-Smoker", "Smoker"), col = c("blue", "red"), lty = 1)
```

**Interpretation**: The survival curves suggest that mothers who smoke tend to stop breastfeeding earlier than non-smokers.

# Cox Proportional Hazards Model
Next, we fit a Cox proportional hazards model to estimate relative risks.

```{r cox-model}
fit_cox <- coxph(Surv(duration, delta) ~ as.factor(race) + poverty + smoke + 
                 alcohol + agemth + ybirth + yschool + pc3mth, data = bfeed)
summary(fit_cox)
```

**Interpretation**:
- Smoking increases the hazard of stopping breastfeeding (`HR = 1.28`, p < 0.01).
- Higher maternal education (`yschool`) is associated with longer breastfeeding duration.

# Parametric Accelerated Failure Time (AFT) Models
We compare Weibull, log-logistic, and log-normal AFT models.

```{r aft-models}
fit_weibull <- survreg(Surv(duration, delta) ~ as.factor(race) + poverty + smoke + 
                       alcohol + agemth + ybirth + yschool + pc3mth, 
                       data = bfeed, dist = "weibull")
fit_loglogistic <- survreg(Surv(duration, delta) ~ as.factor(race) + poverty + smoke + 
                           alcohol + agemth + ybirth + yschool + pc3mth, 
                           data = bfeed, dist = "loglogistic")
fit_lognormal <- survreg(Surv(duration, delta) ~ as.factor(race) + poverty + smoke + 
                          alcohol + agemth + ybirth + yschool + pc3mth, 
                          data = bfeed, dist = "lognormal")

list(weibull = logLik(fit_weibull), loglogistic = logLik(fit_loglogistic), lognormal = logLik(fit_lognormal))
```

**Interpretation**:
- The log-normal model provides the best fit based on log-likelihood values.
- Results are consistent with the Cox model, confirming smoking negatively impacts breastfeeding duration.

# Conclusion
Survival analysis of breastfeeding duration shows that maternal smoking significantly shortens breastfeeding duration. Higher education is associated with prolonged breastfeeding. The log-normal AFT model best fits the data, supporting findings from the Cox model.

This analysis highlights the importance of maternal characteristics in influencing breastfeeding behavior and can inform public health policies.

