---
title: "Advanced Survival Analysis: A Comprehensive Study of Cure Models"
author: "Andrew Kamya"
date: "December, 2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Survival analysis is a cornerstone of statistical methods used to analyze time-to-event data, particularly in medical research. Traditional survival models assume that all individuals in the population are susceptible to the event of interest (e.g., death, relapse). However, in many real-world scenarios, a subset of individuals may be "cured" or no longer at risk of the event. Cure models extend traditional survival analysis by incorporating the possibility of a cured fraction, making them particularly useful in cancer research and other fields where long-term survival is observed.

This report explores advanced topics in survival analysis, focusing on **cure models**. We analyze three datasets using parametric and semiparametric cure models, evaluate their performance, and interpret the results. The analysis is divided into three parts, each addressing a specific dataset and research question.

---

# Methodology

We employ both parametric  and  semiparametric  cure models to analyze survival data. Parametric models assume specific distributions for the survival times (e.g., Weibull, Gamma, Log-normal), while semiparametric models, such as the proportional hazards (PH) mixture cure model, do not require such assumptions. The models are fitted using R packages such as `flexsurv`, `smcure`, and `survival`.

Key steps in the analysis include:
1. Data Preparation: Loading and exploring the datasets.
2. Model Fitting: Fitting parametric and semiparametric cure models.
3. Model Selection: Using AIC to select the best-fitting model.
4. Interpretation: Analyzing the coefficients and cure fractions.
5. Visualization: Plotting survival curves and predicted cure rates.

---

# Part I: Breast Cancer Survival Data Analysis

## Data Description
The `bc` dataset, available in the `flexsurv` package, contains survival times of 686 patients with primary node-positive breast cancer. The variables include:
- `censrec`: Event indicator (1 = dead, 0 = censored).
- `rectime`: Time of death or censoring in days.
- `group`: Prognostic group (Good, Medium, Poor).

## Analysis

### Step 1: Exploratory Analysis
We begin by visualizing the survival curves for the entire dataset and by prognostic group.

```{r}
library("flexsurv")
library("flexsurvcure")

data("bc")

s1 <- survfit(Surv(rectime, censrec) ~ 1, data = bc)
plot(s1, mark.time = TRUE, conf.int = FALSE)

s2 <- survfit(Surv(rectime, censrec) ~ group, data = bc)
plot(s2, col = c('black', 'green', 'red'), lty = 1:3, mark.time = TRUE, lwd = 2)
legend('bottomleft', c("Good", "Medium", "Poor"), lty = 1:3, col = c('black', 'green', 'red'))
```

### Step 2: Fitting Parametric Cure Models
We fit parametric cure models with Weibull, Gamma, Exponential, and Log-normal distributions for the latency part and logistic regression for the incidence part.

```{r}
para_bc_gamma <- flexsurvcure(Surv(rectime, censrec) ~ group, data = bc,
                              anc = list(rate = ~group), dist = "gamma", link = "logistic", mixture = TRUE)

para_bc_weibell <- flexsurvcure(Surv(rectime, censrec) ~ group, data = bc,
                                anc = list(scale = ~group), dist = "weibull", link = "logistic", mixture = TRUE)

para_bc_lnorm <- flexsurvcure(Surv(rectime, censrec) ~ group, data = bc,
                              anc = list(meanlog = ~group), dist = "lnorm", link = "logistic", mixture = TRUE)

para_bc_exp <- flexsurvcure(Surv(rectime, censrec) ~ group, data = bc,
                            anc = list(rate = ~group), dist = "exp", link = "logistic", mixture = TRUE)

AIC(para_bc_gamma, para_bc_weibell, para_bc_lnorm, para_bc_exp)
```

### Step 3: Interpretation
The Log-normal model has the lowest AIC and is selected as the best-fitting model. The results indicate:
- Incidence: Prognostic group significantly impacts the probability of being cured.
    - Since the "flexsurvcure" uses logistic regression to model the probability of being cured, odds ratio of exp(-0.6138)=0.541 to be cured for median prognostic group compared to the good prognostic group.
    - OR of exp(-1.585)= 0.205 to be cured for poor prognostic group compared to the good prognostic group.

- Latency: Prognostic group does not significantly affect the survival time of uncured patients (confidence intervals include 0).


- The mean log survival time in median prognostic group is 0.524 days. Shorter than the mean log survival time in good prognostic group.

- The mean log survival time in poor prognostic group is 1.008 days. Shorter than the mean log survival time in good prognostic group.


---

# Part II: Clinical Trial Data Analysis

## Data Description
The `e1684` dataset, available in the `smcure` package, contains data from a clinical trial evaluating high-dose interferon alpha-2b as postoperative adjuvant therapy. The variables include:
- `TRT`: Treatment group (0 = control, 1 = treatment).
- `SEX`: Gender (0 = male, 1 = female).
- `AGE`: Age (centered to the mean).
- `FAILTIME`: Relapse-free survival time in years.
- `FAILCENS`: Event indicator (1 = relapse, 0 = censored).

## Analysis

### Step 1: Kaplan-Meier Estimators
We estimate and plot the overall survival curves and survival curves by treatment and gender.

```{r}
library("survival")
library("smcure")

data(e1684)

E2_s <- survfit(Surv(FAILTIME, FAILCENS) ~ 1, data = e1684)
plot(E2_s, mark.time = TRUE, conf.int = FALSE)

E2_treat <- survfit(Surv(FAILTIME, FAILCENS) ~ TRT, data = e1684)
plot(E2_treat, col = c('black', 'red'), lty = 1:2, mark.time = TRUE, lwd = 2)
legend('bottomleft', c("Control", "Treatment"), lty = 1:2, col = c('black', 'red'))
```
- Patients in the treatment group have higher cure rate than those in control group.
- Survival curves show that the difference in the cure rates between male and female patients is not significant.
---

### Step 2: Semiparametric PH Mixture Cure Model
We fit a semiparametric proportional hazards mixture cure model using treatment, gender, and age as covariates.

```{r}
sm.ph <- smcure(Surv(FAILTIME, FAILCENS) ~ TRT + SEX + AGE, cureform = ~TRT + SEX + AGE, data = e1684, model = "ph")
printsmcure(sm.ph)
```

### Step 3: Bootstrap Variance Estimation
We estimate the variance of the coefficients using bootstrap resampling.

```{r}
sm.ph100 <- smcure(Surv(FAILTIME, FAILCENS) ~ TRT + SEX + AGE, cureform = ~TRT + SEX + AGE, data = e1684, model = "ph", nboot = 100)
sm.ph200 <- smcure(Surv(FAILTIME, FAILCENS) ~ TRT + SEX + AGE, cureform = ~TRT + SEX + AGE, data = e1684, model = "ph", nboot = 200)
sm.ph500 <- smcure(Surv(FAILTIME, FAILCENS) ~ TRT + SEX + AGE, cureform = ~TRT + SEX + AGE, data = e1684, model = "ph", nboot = 500)
printsmcure(sm.ph100); printsmcure(sm.ph200); printsmcure(sm.ph500)
```

---

# Part III: Bone Marrow Transplant Data Analysis

## Data Description
The `bmt` dataset, available in the `smcure` package, contains data from a bone marrow transplant study. The variables include:
- `TRT`: Treatment group (0 = allogeneic, 1 = autologous).
- `Time`: Time to death.
- `Status`: Event indicator (1 = death, 0 = censored).

## Analysis

### Step 1: Kaplan-Meier Estimators
We estimate and plot the overall survival curves and survival curves by treatment group.

```{r}
data(bmt, package = "smcure")

E3_s1 <- survfit(Surv(Time, Status) ~ 1, data = bmt)
plot(E3_s1, mark.time = TRUE, conf.int = FALSE)

E3_s2 <- survfit(Surv(Time, Status) ~ TRT, data = bmt)
plot(E3_s2, col = c('black', 'red'), lty = 1:2, mark.time = TRUE, lwd = 2)
legend('bottomleft', c("Allogeneic", "Autologous"), lty = 1:2, col = c('black', 'red'))
```

### Step 2: Semiparametric PH and AFT Mixture Cure Models
We fit semiparametric PH and AFT mixture cure models.

```{r}
sm.ph <- smcure(Surv(Time, Status) ~ TRT, cureform = ~TRT, data = bmt, model = "ph", Var = TRUE)
# sm.aft <- smcure(Surv(Time, Status) ~ TRT, cureform = ~TRT, data = bmt, model = "aft", Var = FALSE)
```

---

# Conclusion

This report demonstrates the application of cure models in survival analysis using three distinct datasets. Key findings include:
1. Breast Cancer Data: The Log-normal cure model provided the best fit, with prognostic group significantly impacting the cure probability.
2. Clinical Trial Data: Treatment and gender did not significantly affect the cure probability or survival time.
3. Bone Marrow Transplant Data: The allogeneic treatment group showed a higher cure rate compared to the autologous group.

Cure models are powerful tools for analyzing survival data with a cured fraction, providing insights that traditional survival models cannot capture. Future work could explore more complex models and larger datasets to further validate these findings.

---